---
title: "Week 5: Profiling and Parallel Processing"
output:
  html_document:
    df_print: paged
  github_document: default
---


# Objectives  
- Define functional programming, scoping, and applications.  
- Profiling and parallel computing.  


# Data 


# Functional programming  


Functions are directives in R to simplify repeatable tasks or error-prone tasks.  

```{r, message=FALSE}

round(0.5)
round(1.5)
round(2.5)
round(3.5)

# R uses a banker's rounding form where 0.5 rounds to nearest even integer.

# very common example from stackoverflow
# https://stackoverflow.com/questions/12688717/round-up-from-5

round2 = function(x, n) {
  posneg = sign(x)
  z = abs(x)*10^n
  z = z + 0.5 + sqrt(.Machine$double.eps)
  z = trunc(z)
  z = z/10^n
  z*posneg
}

# round2 allows us to remedy this

round2(0.5,0)
round2(1.5,0)
round2(2.5,0)
round2(3.5,0)

# note we need to use the second parameter. round() defaults to digits=0, but round2() does not.

round2 = function(x, n=0) {
  posneg = sign(x)
  z = abs(x)*10^n
  z = z + 0.5 + sqrt(.Machine$double.eps)
  z = trunc(z)
  z = z/10^n
  z*posneg
}

round2(0.5)
round2(1.5)

```


Functions require you to be explicit when printing.  

```{r, message=FALSE}
x = 1
y = 1
z = x + y
x
y
z

tst <- function(){
x = 1
y = 1
z = x + y
x
y
z
}
tst()
```


Other random function things.  

```{r, message=FALSE}
# can call arguments
args(round)

# partial matching of arguments is allowed and after argument called, it goes remaining positional matching
round(dig=2,5.7777)
```


You can build on R functions using the `...` argument.  

```{r, message=FALSE}

round2 = function(x, n=0) {
  posneg = sign(x)
  z = abs(x)*10^n
  z = z + 0.5 + sqrt(.Machine$double.eps)
  z = trunc(z)
  z = z/10^n
  z*posneg
}

# this means the user needs to provide at least x, but any other parameters are optional but will be passed to the function appropriately
round2perc = function(x, ...){
  x = round2(x*100, ...)
  paste0(as.character(x),"%")
}

# round2 as normal
round2(0.9575,2)

# round2perc with no "n" defined
round2perc(0.9575)

# round2perc with optional "n" defined
round2perc(0.9575,3)

```


You can use functions in functions.  

```{r, message=FALSE}

# declaring function in function
make.power <- function(n) {
  pow <- function(x) {
    x^n
    }
  pow 
}

# we are passing n to the parent function
cube <- make.power(3)
square <- make.power(2)

# we are passing x to the two functions with 'n' defined differently
cube(3)
square(3)
```

Note: be careful with doing this. There can be some odd scoping rules.

# Profiling

Profiling can be a good way to see why things are moving slowly. Expert R programmers will use this to determine what routines need to be sped up and to determine where bottlenecks are in workflows.

`system.time` is a simple routine to see how long something takes.

```{r, message=FALSE}

# homepage of wikipedia
system.time(readLines("https://en.wikipedia.org"))

# longest wikipedia page as of Nov 2021
system.time(readLines("https://en.wikipedia.org/wiki/Border_control"))

```

`Rprof` and `summaryRprof` can be used to interpret how long subroutines take

```{r, message=FALSE}

Rprof() # start Rprof
system.time(readLines("https://en.wikipedia.org/wiki/Border_control"))
Rprof(NULL) # end Rprof
summaryRprof() # looks at profiler output

```

# Parallel computing

Some functions, like crossprod() and basic mathematics (Basic Linear Algebra Subroutines - BLAS) are already parallelized for you.  

## Resources available

```{r,message=FALSE}
library(parallel)

detectCores() # means I have 12 processors available to me

```

```{r}
library(parallel)

#r <- mclapply(1:10, function(i) {
#  Sys.sleep(2)  ## Do nothing for 10 seconds
#  }, mc.cores = 10)      ## Split this job across 10 cores
# r
```

# Resources

- [R Programming for Data Science: Functions](https://bookdown.org/rdpeng/rprogdatascience/functions.html)
- [R Programming for Data Science: Scoping](https://bookdown.org/rdpeng/rprogdatascience/scoping-rules-of-r.html)
- [R Programming for Data Science: Profiling R Code](https://bookdown.org/rdpeng/rprogdatascience/profiling-r-code.html)  
- [R Programming for Data Science: Parallel Computing](https://bookdown.org/rdpeng/rprogdatascience/parallel-computation.html)  