---
title: "Week 4: Optimization & Tidymodels"
output:
  html_document:
    df_print: paged
  github_document: default
---


# Objectives  

- Experimenting with Tidymodels methods.  
- Experimenting with Optimization techniques and case studies.  

# Data 

Data used is taken from [Tidymodels - Get Started](https://www.tidymodels.org/start/models/)  

```{r, message=FALSE}
library(tidyverse)
library(tidymodels)
library(dotwhisker)
library(nycflights13) 
```

# Linear regression example

The example dataset used explores whether the size of sea urchin is impacted by each of the three different food regimes.

```{r, message=FALSE}
urchins <-
  # Data were assembled for a tutorial 
  # at https://www.flutterbys.com.au/stats/tut/tut7.5a.html
  read_csv("https://tidymodels.org/start/models/urchins.csv") %>% 
  # Change the names to be a little more verbose
  setNames(c("food_regime", "initial_volume", "width")) %>% 
  # Factors are very helpful for modeling, so we convert one column
  mutate(food_regime = factor(food_regime, levels = c("Initial", "Low", "High")))

urchins %>% sample_n(20) %>% head(20)
```

A plot of each urchin and its food_regime

```{r, message=FALSE}
ggplot(urchins,
       aes(x = initial_volume, 
           y = width, 
           group = food_regime, 
           col = food_regime)) + 
  geom_point() + 
  geom_smooth(method = lm, se = FALSE) +
  scale_color_viridis_d(option = "plasma", end = .7) +
  ggtitle("Sea urchin by their food regime")
```

Define a model for tidymodels to use using {{parsnip}} of {{tidymodels}} functions (linear_reg, set_engine, fit)

```{r, message=FALSE}
# linear regression model with the engine set as the ordinary least squares ('lm') package
lm_mod <- 
  linear_reg() %>% 
  set_engine("lm")

# apply the modeling equation
lm_fit <- 
  lm_mod %>% 
  fit(width ~ initial_volume * food_regime, data = urchins)

tidy(lm_fit)
```

Tidymodels allows you to generate a dot-and-whisker plot of the regression results.

```{r, message=FALSE}
tidy(lm_fit) %>% 
  dwplot(dot_args = list(size = 2, color = "black"),
         whisker_args = list(color = "black"),
         vline = geom_vline(xintercept = 0, colour = "grey50", linetype = 2))
```

From there, you can predict across new points against the former model.

```{r, message=FALSE}
new_points <- expand.grid(initial_volume = 20, 
                          food_regime = c("Initial", "Low", "High"))
new_points

mean_pred <- predict(lm_fit, new_data = new_points)
mean_pred
```

You can also ask predict() to provide confidence intervals, and all of the data can be easily combined across both the mean prediction and CI prediction.  
```{r, message=FALSE}
conf_int_pred <- predict(lm_fit, 
                         new_data = new_points, 
                         type = "conf_int")
conf_int_pred

# Now combine: 
plot_data <- 
  new_points %>% 
  bind_cols(mean_pred) %>% 
  bind_cols(conf_int_pred)

# and plot:
ggplot(plot_data, aes(x = food_regime)) + 
  geom_point(aes(y = .pred)) + 
  geom_errorbar(aes(ymin = .pred_lower, 
                    ymax = .pred_upper),
                width = .2) + 
  labs(y = "urchin size") + ggtitle("Prediction of 3 urchins of initial volume of 20 across 3 types of food_regime")
```

# Logistic regression example

Note a logistic regression dataset should have a categorical variable (1/0, yes/no, test/fail) as the dependent variable (what you are predicting).  

```{r, message=FALSE}
flight_data <- 
  flights %>% 
  mutate(
    # Convert the arrival delay to a factor
    arr_delay = ifelse(arr_delay >= 30, "late", "on_time"),
    arr_delay = factor(arr_delay),
    # We will use the date (not date-time) in the recipe below
    date = lubridate::as_date(time_hour)
  ) %>% 
  # Include the weather data
  inner_join(weather, by = c("origin", "time_hour")) %>% 
  # Only retain the specific columns we will use
  select(dep_time, flight, origin, dest, air_time, distance, 
         carrier, date, arr_delay, time_hour) %>% 
  # Exclude missing data
  na.omit() %>% 
  # For creating models, it is better to have qualitative columns
  # encoded as factors (instead of character strings)
  mutate_if(is.character, as.factor)

head(flight_data,20)
```

Note that 16% of flights arrive late.

```{r, message=FALSE}
flight_data %>% 
  count(arr_delay) %>% 
  mutate(prop = n/sum(n))
```

Splitting data into test and training set using {{rsample}} of {{tidymodels}}.

```{r,message=FALSE}
# Fix the random numbers by setting the seed 
# This enables the analysis to be reproducible when random numbers are used 
set.seed(222)
# Put 3/4 of the data into the training set 
data_split <- initial_split(flight_data, prop = 3/4)

# Create data frames for the two sets:
train_data <- training(data_split)
test_data  <- testing(data_split)
```

Create a recipe and roles in tidymodels.

```{r, message=FALSE}
# assign a recipe where all variables are predictors of arr_delay
flights_rec <- 
  recipe(arr_delay ~ ., data = train_data) 

# assign two variables as "ID" (user-defined group). these variables are neither predictors or outcomes
flights_rec <- 
  recipe(arr_delay ~ ., data = train_data) %>% 
  update_role(flight, time_hour, new_role = "ID") 

summary(flights_rec)
```

Create a more refined recipe with mutated date objects using {{recipes}} functions 'step_date' and 'step_holiday'

```{r, message=FALSE}
# new recipe
flights_rec <- 
  recipe(arr_delay ~ ., data = train_data) %>% 
  update_role(flight, time_hour, new_role = "ID") %>% 
  # output a feature for day of week (dow) and for month
  step_date(date, features = c("dow", "month")) %>%     
  # output a feature for holiday (yes/no) and remove 'date' as a feature in the dataset
  step_holiday(date, 
               holidays = timeDate::listHolidays("US"), 
               keep_original_cols = FALSE)
```

Use 'step_dummy' to turn factors like 'dest' and 'origin' columns into indicator/dummy variables.

```{r, message=FALSE}
flights_rec <- 
  recipe(arr_delay ~ ., data = train_data) %>% 
  update_role(flight, time_hour, new_role = "ID") %>% 
  step_date(date, features = c("dow", "month")) %>%               
  step_holiday(date, 
               holidays = timeDate::listHolidays("US"), 
               keep_original_cols = FALSE) %>% 
  # apply across all predictors with character columns
  step_dummy(all_nominal_predictors())
```

Test for variables that are in testing dataset, but not training dataset. This means we will have to exclude them using step_zv() (zero-variance predictor).

```{r, message=FALSE}
test_data %>% 
  distinct(dest) %>% 
  anti_join(train_data)

flights_rec <- 
  recipe(arr_delay ~ ., data = train_data) %>% 
  update_role(flight, time_hour, new_role = "ID") %>% 
  step_date(date, features = c("dow", "month")) %>%               
  step_holiday(date, 
               holidays = timeDate::listHolidays("US"), 
               keep_original_cols = FALSE) %>% 
  step_dummy(all_nominal_predictors()) %>% 
  # exclude any cases where we may not have zero-variance predictors, or where the training set have a single value
  step_zv(all_predictors())
```

Three steps to running a recipe: (1) process the recipe using the training set, (2) apply the recipe to the training set, (3) apply the recipe to the test set.

Create a workflow using {{worfklows}} from {{tidymodels}} to pair together the model and the recipe together. This will make it easier to 

```{r, message=FALSE}
lr_mod <- 
  logistic_reg() %>% 
  set_engine("glm")

flights_wflow <- 
  workflow() %>% 
  add_model(lr_mod) %>% 
  add_recipe(flights_rec)
```

Run the workflow and evaluate the fit.

```{r, message=FALSE}
flights_fit <- 
  flights_wflow %>% 
  fit(data = train_data)

flights_fit %>% 
  extract_fit_parsnip() %>% 
  tidy()
```

Apply the fitted model to the test_data.

```{r, message=FALSE}
# this returns the outcome variable as a factor
predict(flights_fit, test_data)

# we can use an augment function to output predicted class probabilities
flights_aug <- 
  augment(flights_fit, test_data)

flights_aug %>%
  select(arr_delay, time_hour, flight, .pred_class, .pred_on_time)

```

Use a ROC curve to evaluate the model using the {{yardstick}} package of {{tidymodels}}.

```{r, message=FALSE}
flights_aug %>% 
  roc_curve(truth = arr_delay, .pred_late) %>% 
  autoplot()

# estimate the area under the curve
flights_aug %>% 
  roc_auc(truth = arr_delay, .pred_late)
```
