---
title: 'Week 2: EDA and Simple Modelling'
output:
  html_document:
    df_print: paged
  github_document: default
---

# Objectives:

- Exploratory Data Analysis (EDA)
- Simple modelling (linear regression, residuals, logistic regression)

# Exploratory Data Analysis

Exploratory Data Analysis is about:
- univariate non-graphical EDA
- univariate graphical EDA
- multivariate non-graphical EDA
- multivariate graphical EDA

Methods to explore:  
- Variation (describe behavior within variable)  
- Covariation (describe behavior between variables)  
- Missing data / anomalies    
- Patterns / correlations  

Tools:  
- tables  
- scatterplot  
- histogram  
- boxplot  
- heatplot and hex plots  

Pull in new dataset.

Dataset from [Department of Labor](https://www.dol.gov/agencies/eta/foreign-labor/performance) about applications for permanent employment certification. Download the "PERM" FY 2021 Disclosure file (xlsx) and put in folder /lessons/data.

```{r, message=FALSE}
library(tidyverse)
library(readxl)
library(gt)
library(gtExtras) #remotes::install_github("jthomasmock/gtExtras")
# in some cases, we need to change guess_max because the column may have nulls for a lot of the first rows
df = read_excel("lessons/data/PERM_Disclosure_Data_FY2021_Q3.xlsx",guess_max=15000)
head(df,2) %>% as.data.frame()
```

Do general data checks to understand the data you are dealing with.

```{r, message=FALSE}

# validate if all cases are distinct
df %>% distinct(CASE_NUMBER) %>% nrow() == nrow(df)

# look at received yr
df %>% mutate(received_yr=lubridate::year(RECEIVED_DATE)) %>% group_by(received_yr) %>% count() %>% ungroup()

# look at decision yr
df %>% mutate(decision_yr=lubridate::year(DECISION_DATE)) %>% group_by(decision_yr) %>% count() %>% ungroup()

# look at employers
df %>% group_by(EMPLOYER_NAME) %>% count() %>% ungroup() %>% arrange(desc(n))
```

Now, we can make some interesting graphics / tables to describe trends.

```{r, message=FALSE}
summary(df$PW_WAGE)

wages = df %>%
    group_by(PW_WAGE) %>% count() %>% ungroup()

wages %>%
    ggplot(aes(x=PW_WAGE,y=n)) + geom_bar(stat='identity',width=20) + theme_bw()

wages %>%
  ggplot(aes(x=PW_WAGE,y=n)) + geom_point() + theme_bw()

df %>%
  group_by(PW_WAGE) %>% count() %>% ungroup() %>%
  ggplot(aes(x=PW_WAGE,y=n)) + geom_point() + theme_bw()

df %>%
  group_by(PW_WAGE) %>% count() %>% ungroup() %>%
  ggplot(aes(x=PW_WAGE,y=n)) + geom_point() + theme_bw()

top_10_employers = df %>% group_by(EMPLOYER_NAME) %>% count() %>% ungroup() %>% arrange(desc(n)) %>% head(10)

df_top_10 = df %>% dplyr::filter(EMPLOYER_NAME %in% top_10_employers$EMPLOYER_NAME)

gg_all = df %>% 
  ggplot(aes(x=PW_WAGE)) +
  geom_histogram(breaks=seq(0,300000,by=10000),fill="black",colour="white") +
  scale_x_continuous(name="Wage",breaks=seq(0,300000,by=50000),labels=scales::comma) +
  scale_y_continuous(name="Applicants w/ Wage",labels=scales::comma) +
  theme_bw() +
  ggtitle("Wages per Applicant (2020-2021)")

gg_all

gg_top10 = df %>% 
  dplyr::filter(EMPLOYER_NAME %in% top_10_employers$EMPLOYER_NAME) %>%
  ggplot(aes(x=PW_WAGE)) +
  geom_histogram(breaks=seq(0,300000,by=10000),fill="black",colour="white") +
  scale_x_continuous(name="Wage",breaks=seq(0,300000,by=50000),labels=scales::comma) +
  scale_y_continuous(name="Applicants w/ Wage",labels=scales::comma) +
  theme_bw() +
  ggtitle("Wages per Applicant (2020-2021)")
  
ggplot() +
  geom_histogram(data = df,mapping=aes(x=PW_WAGE),breaks=seq(0,300000,by=10000),fill="grey",colour="white") +
  geom_histogram(data = df_top_10,mapping=aes(x=PW_WAGE),breaks=seq(0,300000,by=10000),fill="black",colour="white") +
  scale_x_continuous(name="Wage",breaks=seq(0,300000,by=50000),labels=scales::comma) +
  scale_y_continuous(name="Applicants w/ Wage",labels=scales::comma) +
  theme_bw() +
  ggtitle(label = "Wages per Applicant (2020-2021)", subtitle = "Black represents top 10 employers")

df_top_10 = df %>% dplyr::filter(EMPLOYER_NAME %in% top_10_employers$EMPLOYER_NAME)

top_paid = df %>% arrange(desc(PW_WAGE)) %>% top_frac(0.1,wt = PW_WAGE) %>% 
  group_by(EMPLOYER_NAME) %>% summarise(n=n(),sum=sum(PW_WAGE,na.rm=TRUE),max=max(PW_WAGE,na.rm=TRUE),med=stats::median(PW_WAGE,na.rm=TRUE)) %>% ungroup() %>% 
  arrange(desc(n)) %>% head(10)

df %>% dplyr::filter(EMPLOYER_NAME %in% top_paid$EMPLOYER_NAME)

# gtExtras and gt_plt_bullet() is a new package/function and the syntax may change in the future
top_paid %>% 
  dplyr::mutate(plot_column = max) %>%
  gt() %>% gt_plt_bullet(column=plot_column,target=med,width=45,colors= c("lightblue","black")) %>%
  gt::cols_label(EMPLOYER_NAME=md("Employer"),n=md("Count"),sum=md("Sum"),max=md("Wage")) %>% gt::fmt_number(columns=c(sum,max),decimals=0)

```

# Resources

## Exploratory Data Analysis (EDA)  

- [R for Data Science: Exploratory Data Analysis](https://r4ds.had.co.nz/exploratory-data-analysis.html)  
- [StackExchange on EDA](https://stats.stackexchange.com/questions/114610/what-is-the-relationship-between-y-and-x-in-this-plot)  
- [Experimental Design and Analysis by Howard J. Seltman (2018), Chapter 4: Exploratory Data Analysis](https://www.stat.cmu.edu/~hseltman/309/Book/chapter4.pdf)  
- [Exploratory Data Analysis - EPA](https://www.epa.gov/caddis-vol4/exploratory-data-analysis)  
- [Biomedical Data Science, Exploratory Data Analysis](https://genomicsclass.github.io/book/pages/exploratory_data_analysis.html)
- [Beginners Guide to Explanatory Data Analysis by Irem Tanrıverdi](https://www.analyticsvidhya.com/blog/2021/04/beginners-guide-to-explanatory-data-analysis/)
